### 2021-12-06 lmcs/priority-gv.tex
work on pgv

\subsubsection*{Session types}
Session types ($\ty{S}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \Coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
Session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. Types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

\subsubsection*{Types}
Types ($\ty{T}$, $\ty{U}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \Coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$.

\subsubsection*{Environments}
Typing environments $\ty{\Gamma}$, $\ty{\Delta}$ associate types to names. Environments are linear, so two environments can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if their names are distinct, \ie $\fv(\ty{\Gamma})\cap\fv(\ty{\Delta})=\varnothing$.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \Coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]

\subsubsection*{Duality}
Duality plays a crucial role in session types. The two endpoints of a channel are assigned dual types, ensuring that, for instance, whenever one program \emph{sends} a value on a channel, the program on the other end is waiting to \emph{receive}. Each session type $\ty{S}$ has a dual, written $\ty{\co{S}}$. Duality is an involutive function which \emph{preserves priorities}:
\[
  \ty{\co{\tysend[\cs{o}]{T}{S}}} = \ty{\tyrecv[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} = \ty{\tysend[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyends[\cs{o}]}} = \ty{\tyendr[\cs{o}]}
  \qquad
  \ty{\co{\tyendr[\cs{o}]}} = \ty{\tyends[\cs{o}]}
\]

@@

\subsubsection*{Internal and External Choice}
Typically, session-typed languages feature constructs for internal and external choice. In GV, these can be defined in terms of the core language, by sending or receiving a value of a sum type~\cite{lindleymorris15}. We use the following syntactic sugar for internal ($\ty{\tyselect[\cs{o}]{S}{S'}}$) and external ($\ty{\tyoffer[\cs{o}]{S}{S'}}$) choice and their units:
\[
\begin{array}{lcl}
  \ty{\tyselect[\cs{o}]{S}{S'}}
  & \elabarrow & \ty{\tysend[\cs{o}]{(\tysum{\co{S}}{\co{S'}})}{\tyends[\cs{o+1}]}} \\
  \ty{\tyoffer[\cs{o}]{S}{S'}}
  & \elabarrow & \ty{\tyrecv[\cs{o}]{(\tysum{S}{S'})}{\tyendr[\cs{o+1}]}} \\
\end{array}
\qquad
\begin{array}{lcl}
  \ty{\tyselectemp[\cs{o}]}
  & \elabarrow & \ty{\tysend[\cs{o}]{\tyvoid}{\tyends[\cs{o+1}]}} \\
  \ty{\tyofferemp[\cs{o}]}
  & \elabarrow & \ty{\tyrecv[\cs{o}]{\tyvoid}{\tyendr[\cs{o+1}]}} \\
\end{array}
\]
As the syntax for units suggests, these are the binary and nullary forms of the more common n-ary choice constructs $\ty{{\oplus}^{\cs{o}}\{l_i:S_i\}_{i\in{I}}}$ and $\ty{{\with}^{\cs{o}}\{l_i:S_i\}_{i\in{I}}}$, which one may obtain generalising the sum types to variant types. For simplicity, we present only the binary and nullary forms.

@@

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
  \tm{\phi}
  \Coloneqq \tm{\main}
  \sep      \tm{\child}
  \hfill
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  \Coloneqq \tm{\phi\;M}
  \sep      \tm{\ppar{\conf{C}}{\conf{D}}}
  \sep      \tm{\res{x}{x'}{\conf{C}}}
  \hfill
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \hfill
  \tm{\child} + \tm{\main}  = \tm{\main}
  \hfill
  \tm{\child} + \tm{\child} = \tm{\child}
  \hfill
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
  \hfill
\]
The use of $\tm{\child}$ for child threads~\cite{lindleymorris15} overlaps with the use of the meta-variable $\cs{o}$ for priorities~\cite{dardhagay18extended}. Both are used to annotate sequents: flags appear on the sequent in configuration typing, and priorities in term typing. To distinguish the two symbols, they are typeset in a different font and a different colour.

Values ($\tm{V}$, $\tm{W}$), evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{V}{W}}
    \sep        \tm{\inl{V}}
    \sep        \tm{\inr{V}} \\
  \tm{E}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\andthen{E}{N}}
    \sep        \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}}
    \sep        \tm{\absurd{E}} \\
  \tm{\conf{F}}
  & \Coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]

\input{fig/pgv-operational-semantics}

@@

\subsubsection*{Typing}
\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}. Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat \emph{all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$. However, the rules can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

\input{fig/pgv-typing}

@@

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well-typed, they impose opposite conditions on the priorities, so their composition is ill-typed. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \small
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

  \inferrule*{
    \tseq[\cs{o}]
    {\tmty{x'}{\tyrecv[\cs{o}]{\tyunit}{\tyendr}}}
    {\recv\;{x'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{q}]
    {\tmty{y}{\tysend[\cs{o'}]{\tyunit}{\tyends}},\tmty{x'}{\tyendr}}
    {\letbind{y}{\send\;{\pair{\unit}{y}}}{\dots}}
    {\tyunit}
    \and
    \cs{o}<\cs{o'}
  }{%
    \tseq[\cs{q}]
    {%
      \tmty{y}{\tysend[\cs{o'}]{\tyunit}{\tyends}},
      \tmty{x'}{\tyrecv[\cs{o}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{x'}{\recv\;{x'}}{\letbind{y}{\send\;{\pair{\unit}{y}}}{\dots}}}
    {\tyunit}
  }
\end{mathpar}
Closures suspend communication, so \LabTirName{T-Lam} stores the priority bounds of the function body on the function type, and \LabTirName{T-App} restores them. For instance, $\tm{\lambda{x}.\send\;\pair{x}{y}}$ is assigned the type $\ty{\tylolli[\cs{o},\cs{o}]{A}{S}}$, \ie, a~function which, when applied, starts and finishes communicating at priority $\cs{o}$.
\begin{mathpar}
  \small
  \inferrule*{
    \inferrule*{
      \inferrule*{
      }{
        \tmty{\send}{\tylolli[\ptop,\cs{o}]{\typrod{A}{\tysend[\cs{o}]{A}{S}}}{S}}
      }
      \and
      \inferrule*{%
        \inferrule*{
        }{
          \tseq[\pbot]{\tmty{x}{A}}{x}{A}
        }
        \and
        \inferrule*{
        }{
          \tseq[\pbot]
          {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
          {y}{\tysend[\cs{o}]{A}{S}}
        }
      }{
        \tseq[\pbot]
        {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
        {\pair{x}{y}}{\typrod{A}{\tysend[\cs{o}]{A}{S}}}
      }
    }{
      \tseq[\cs{o}]
      {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
      {\send\;\pair{x}{y}}{S}
    }
  }{
    \tseq[\pbot]
    {\tmty{y}{\tysend[\cs{o}]{A}{S}}}
    {\lambda{x}.\send\;\pair{x}{y}}{\tylolli[\cs{o},\cs{o}]{A}{S}}
  }
\end{mathpar}

Typing rules for configurations are at the bottom of \cref{fig:pgv-typing}. Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ stating that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensures that there is at most one main thread.

While our configuration typing is based on the standard typing for GV, it differs on two counts:
\begin{enumerate*}[label= (\roman*) ]
\item \emph{we require that child threads return the unit value}, as opposed to a terminated channel; and
\item \emph{we simplify typing for parallel composition}.
\end{enumerate*}
In order to guarantee deadlock freedom, in GV each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$. Consequently, associativity of parallel composition does not preserve typing. In PGV, we guarantee deadlock freedom using priorities, which removes the need for the channel pseudo-type $\ty{S^\sharp}$, and simplifies typing for parallel composition, while restoring type preservation for the structural congruence.

\section{Technical Developments}
\subsection{Subject reduction}
Unlike with previous versions of GV, structural congruence, term reduction, and configuration reduction are all type preserving.

@@

\subsection{Progress and deadlock freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well-typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV. Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged. Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x}$, and then returns $\tm{x'}$:
\[
  \begin{array}{ll}
    \tm{\main}
    & \tm{\letpair{x}{x'}{\new}{}}
    \\
    & \tm{\andthen{\spawn\;(\lambda\unit.\echo_{x})}{x'}}
  \end{array}
  \hfill
  \begin{array}{lcl}
    \tm{\echo_x}
    & \defeq
    & \tm{\letpair{y}{x}{\recv\;{x}}{}}
    \\
    &
    & \tm{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}
  \end{array}
  \hfill
\]
If we reduce the above program, we get $\tm{\res{x}{x'}{(\ppar{\child\;\echo_{x}}{\main\;x'})}}$. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x}}$, which is blocked on $\tm{x}$. In~\cref{cor:pgv-closed-progress} we will show that if a term does not return an endpoint, it must produce \emph{only} a value.

>>>>

\subsection{Syntax of Types and Terms}
\subsubsection*{Session types}
Session types ($\ty{S}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \Coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
Session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. Types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

\subsubsection*{Regular Types}
Types ($\ty{T}$, $\ty{U}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \Coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$.

\subsubsection*{Typing Environments}
Typing environments $\ty{\Gamma}$, $\ty{\Delta}$ associate types to names. Environments are linear, so two environments can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if their names are distinct, \ie $\fv(\ty{\Gamma})\cap\fv(\ty{\Delta})=\varnothing$.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \Coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]

\subsubsection*{Type Duality}
Duality plays a crucial role in session types. The two endpoints of a channel are assigned dual types, ensuring that, for instance, whenever one program \emph{sends} a value on a channel, the program on the other end is waiting to \emph{receive}. Each session type $\ty{S}$ has a dual, written $\ty{\co{S}}$. Duality is an involutive function which \emph{preserves priorities}:
\[
  \ty{\co{\tysend[\cs{o}]{T}{S}}} = \ty{\tyrecv[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} = \ty{\tysend[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyends[\cs{o}]}} = \ty{\tyendr[\cs{o}]}
  \qquad
  \ty{\co{\tyendr[\cs{o}]}} = \ty{\tyends[\cs{o}]}
\]

@@

\subsubsection*{Internal and External Choice}
Typically, session-typed languages feature constructs for internal and external choice. In GV, these can be defined in terms of the core language, by sending or receiving a value of a sum type~\cite{lindleymorris15}. We use the following syntactic sugar for internal ($\ty{\tyselect[\cs{o}]{S}{S'}}$) and external ($\ty{\tyoffer[\cs{o}]{S}{S'}}$) choice and their units:
\[
\begin{array}{lcl}
  \ty{\tyselect[\cs{o}]{S}{S'}}
  & \elabarrow & \ty{\tysend[\cs{o}]{(\tysum{\co{S}}{\co{S'}})}{\tyends[\cs{o+1}]}} \\
  \ty{\tyoffer[\cs{o}]{S}{S'}}
  & \elabarrow & \ty{\tyrecv[\cs{o}]{(\tysum{S}{S'})}{\tyendr[\cs{o+1}]}} \\
\end{array}
\qquad
\begin{array}{lcl}
  \ty{\tyselectemp[\cs{o}]}
  & \elabarrow & \ty{\tysend[\cs{o}]{\tyvoid}{\tyends[\cs{o+1}]}} \\
  \ty{\tyofferemp[\cs{o}]}
  & \elabarrow & \ty{\tyrecv[\cs{o}]{\tyvoid}{\tyendr[\cs{o+1}]}} \\
\end{array}
\]

As the syntax for units suggests, these are the binary and nullary forms of the more common n-ary choice constructs $\ty{{\oplus}^{\cs{o}}\{l_i:S_i\}_{i\in{I}}}$ and $\ty{{\with}^{\cs{o}}\{l_i:S_i\}_{i\in{I}}}$, which one may obtain generalising the sum types to variant types. For simplicity, we present only the binary and nullary forms.

@@

\subsection{Operational Semantics}
\subsubsection*{Configurations}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
  \tm{\phi}
  \Coloneqq \tm{\main}
  \sep      \tm{\child}
  \hfill
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  \Coloneqq \tm{\phi\;M}
  \sep      \tm{\ppar{\conf{C}}{\conf{D}}}
  \sep      \tm{\res{x}{x'}{\conf{C}}}
  \hfill
\]

Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \hfill
  \tm{\child} + \tm{\main}  = \tm{\main}
  \hfill
  \tm{\child} + \tm{\child} = \tm{\child}
  \hfill
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
  \hfill
\]

The use of $\tm{\child}$ for child threads~\cite{lindleymorris15} overlaps with the use of the meta-variable $\cs{o}$ for priorities~\cite{dardhagay18extended}. Both are used to annotate sequents: flags appear on the sequent in configuration typing, and priorities in term typing. To distinguish the two symbols, they are typeset in a different font and a different colour.

\subsubsection*{Values}
Values ($\tm{V}$, $\tm{W}$), evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{V}{W}}
    \sep        \tm{\inl{V}}
    \sep        \tm{\inr{V}} \\
  \tm{E}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\andthen{E}{N}}
    \sep        \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}}
    \sep        \tm{\absurd{E}} \\
  \tm{\conf{F}}
  & \Coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]

\subsubsection*{Reduction Relation}
\input{fig/pgv-operational-semantics}

@@

\subsection{Typing Rules}
\input{fig/pgv-typing}

\subsubsection*{Terms Typing}
%\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}.
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat \emph{all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$. However, the rules can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

@@

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well-typed, they impose opposite conditions on the priorities, so their composition is ill-typed. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

  \inferrule*{
    \tseq[\cs{o}]
    {\tmty{x'}{\tyrecv[\cs{o}]{\tyunit}{\tyendr}}}
    {\recv\;{x'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{q}]
    {\tmty{y}{\tysend[\cs{o'}]{\tyunit}{\tyends}},\tmty{x'}{\tyendr}}
    {\letbind{y}{\send\;{\pair{\unit}{y}}}{\dots}}
    {\tyunit}
    \and
    \cs{o}<\cs{o'}
  }{%
    \tseq[\cs{q}]
    {%
      \tmty{y}{\tysend[\cs{o'}]{\tyunit}{\tyends}},
      \tmty{x'}{\tyrecv[\cs{o}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{x'}{\recv\;{x'}}{\letbind{y}{\send\;{\pair{\unit}{y}}}{\dots}}}
    {\tyunit}
  }
\end{mathpar}

Closures suspend communication, so \LabTirName{T-Lam} stores the priority bounds of the function body on the function type, and \LabTirName{T-App} restores them. For instance, $\tm{\lambda{x}.\send\;\pair{x}{y}}$ is assigned the type $\ty{\tylolli[\cs{o},\cs{o}]{A}{S}}$, \ie, a~function which, when applied, starts and finishes communicating at priority $\cs{o}$.
\begin{mathpar}
  \inferrule*{
    \inferrule*{
      \inferrule*{
      }{
        \tmty{\send}{\tylolli[\ptop,\cs{o}]{\typrod{A}{\tysend[\cs{o}]{A}{S}}}{S}}
      }
      \and
      \inferrule*{%
        \inferrule*{
        }{
          \tseq[\pbot]{\tmty{x}{A}}{x}{A}
        }
        \and
        \inferrule*{
        }{
          \tseq[\pbot]
          {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
          {y}{\tysend[\cs{o}]{A}{S}}
        }
      }{
        \tseq[\pbot]
        {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
        {\pair{x}{y}}{\typrod{A}{\tysend[\cs{o}]{A}{S}}}
      }
    }{
      \tseq[\cs{o}]
      {\tmty{x}{A},\tmty{y}{\tysend[\cs{o}]{A}{S}}}
      {\send\;\pair{x}{y}}{S}
    }
  }{
    \tseq[\pbot]
    {\tmty{y}{\tysend[\cs{o}]{A}{S}}}
    {\lambda{x}.\send\;\pair{x}{y}}{\tylolli[\cs{o},\cs{o}]{A}{S}}
  }
\end{mathpar}

\subsubsection*{Configurations Typing}
Typing rules for configurations are at the bottom of \cref{fig:pgv-typing}. Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ stating that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensures that there is at most one main thread.

While our configuration typing is based on the standard typing for GV, it differs on two counts:
\begin{enumerate}[label= (\roman*) ]
\item \emph{we require that child threads return the unit value}, as opposed to a terminated channel; and
\item \emph{we simplify typing for parallel composition}.
\end{enumerate}
In order to guarantee deadlock freedom, in GV each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$. Consequently, associativity of parallel composition does not preserve typing. In PGV, we guarantee deadlock freedom using priorities, which removes the need for the channel pseudo-type $\ty{S^\sharp}$, and simplifies typing for parallel composition, while restoring type preservation for the structural congruence.

\subsubsection*{Syntactic Sugar Typing}
\input{fig/pgv-typing-sugar}
The following typing rules given in \cref{{fig:pgv-typing-sugar},fig:pgv-typing-sugar-select,fig:pgv-typing-sugar-offer}, cover syntactic sugar typing for PGV.

\section{Technical Developments}
\subsection{Subject Reduction}
Unlike with previous versions of GV, structural congruence, term reduction, and configuration reduction are all type preserving.

@@

\subsection{Progress and Deadlock Freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well-typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV.

Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged. Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x}$, and then returns $\tm{x'}$:
\[
  \begin{array}{ll}
    \tm{\main}
    & \tm{\letpair{x}{x'}{\new}{}}
    \\
    & \tm{\andthen{\spawn\;(\lambda\unit.\echo_{x})}{x'}}
  \end{array}
  \hfill
  \begin{array}{lcl}
    \tm{\echo_x}
    & \defeq
    & \tm{\letpair{y}{x}{\recv\;{x}}{}}
    \\
    &
    & \tm{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}
  \end{array}
  \hfill
\]

If we reduce the above program, we get $\tm{\res{x}{x'}{(\ppar{\child\;\echo_{x}}{\main\;x'})}}$. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x}}$, which is blocked on $\tm{x}$. In~\cref{cor:pgv-closed-progress} we will show that if a term does not return an endpoint, it must produce \emph{only} a value.
